---
title: Modelling
author: Matthew Gordon
date: today
format: 
  html: 
    self-contained: true
    df-print: tibble
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
  message: false
  echo: false
---

```{r, setup, include = FALSE}
pacman::p_load(ggplot2, tidyverse, tidymodels, readr, here, gglm, patchwork, gt, webshot2, katex)

data <- read_csv(here("data", "clean_data.csv"))

glimpse(data)
```

```{r, include = FALSE}
snp_cols <- data |> 
  select(starts_with("rs")) |> 
  colnames()

data <- data |> 
  mutate(
    treatment = as.factor(treatment),
    sex = as.factor(sex),
    across(all_of(snp_cols), as.factor),     
    PR = as.factor(PR)        # PR is binary → classification problem
  )

glimpse(data)
```

```{r, include = FALSE}
set.seed(123)

# Why: split data to evaluate model on unseen data
split_AF <- initial_split(data, strata = treatment)

train_AF <- training(split_AF)
test_AF  <- testing(split_AF)
```

```{r, include = FALSE}
# =========================
# SIMPLE LINEAR REGRESSION
# AF ~ treatment + age
# =========================

# Recipe
recipe_simple <- recipe(AF ~ treatment + age, data = train_AF) |>
  step_dummy(all_nominal_predictors()) |>
  step_center(age)

# Model
model_simple <- linear_reg() |> 
  set_engine("lm")

# Workflow
wf_simple <- workflow(
  recipe_simple,
  model_simple
)

# Fit model
fit_simple <- wf_simple |> 
  fit(train_AF)

fit_simple

fit_simple |> 
  extract_fit_parsnip() |> 
  tidy()

pred_simple <- predict(fit_simple, test_AF) |> 
  bind_cols(test_AF)

pred_vs_actual_simp_lr <- ggplot(pred_simple, aes(AF, .pred, colour = treatment)) +
  geom_point(alpha = 0.7) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Simple Linear Regression",
    x = "Actual AF",
    y = "Predicted AF"
  ) +
  theme_minimal()


plot_data <- expand_grid(
  age = seq(
    min(data$age, na.rm = TRUE),
    max(data$age, na.rm = TRUE),
    length.out = 100
  ),
  treatment = levels(data$treatment)
)

plot_data <- plot_data |> 
  bind_cols(
    predict(fit_simple, new_data = plot_data)
  )

AF_vs_age_simp_lm <- ggplot() +
  geom_point(
    data = data,
    aes(age, AF, colour = treatment),
    alpha = 0.5
  ) +
  geom_line(
    data = plot_data,
    aes(age, .pred, colour = treatment),
    linewidth = 1.2
  ) +
  labs(
    title = "Predicted AF vs Age",
    x = "Age",
    y = "AF"
  ) +
  theme_minimal()
```

```{r}
lm_engine <- fit_simple |> 
  extract_fit_engine()

gglm_simp_lm <- gglm(lm_engine)
```

```{r, include = FALSE}
# =========================
# INTERACTION LINEAR MODEL
# AF ~ treatment * age
# =========================

# Recipe
recipe_interact <- recipe(
  AF ~ treatment + age,
  data = train_AF
) |>
  step_impute_mean(age) |>
  step_center(age) |>
  step_dummy(all_nominal_predictors()) |>
  step_interact(
    terms = ~ age:starts_with("treatment_")
  )

# Model
model_interact <- linear_reg() |> 
  set_engine("lm")

# Workflow
wf_interact <- workflow(
  recipe_interact,
  model_interact
)

# Fit model
fit_interact <- wf_interact |> 
  fit(train_AF)

fit_interact

# =========================
# VIEW COEFFICIENTS
# =========================

fit_interact |> 
  extract_fit_parsnip() |> 
  tidy()

# =========================
# TEST SET PREDICTIONS
# =========================

pred_interact <- predict(fit_interact, test_AF) |> 
  bind_cols(test_AF)

# =========================
# PREDICTED VS ACTUAL AF
# =========================

pred_vs_actual_inter <- ggplot(pred_interact, aes(x = AF, y = .pred, colour = treatment)) +
  geom_point(alpha = 0.7) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Interaction Model",
    x = "Actual AF",
    y = "Predicted AF",
    colour = "Treatment"
  ) +
  theme_minimal()

# =========================
# PREDICTED AF VS AGE
# =========================

plot_interact <- expand_grid(
  age = seq(
    min(data$age, na.rm = TRUE),
    max(data$age, na.rm = TRUE),
    length.out = 200
  ),
  treatment = levels(data$treatment)
)

plot_interact <- plot_interact |> 
  bind_cols(
    predict(fit_interact, new_data = plot_interact)
  )

AF_vs_age_inter <- ggplot() +
  
  geom_point(
    data = data,
    aes(x = age, y = AF, colour = treatment),
    alpha = 0.4
  ) +
  
  geom_line(
    data = plot_interact,
    aes(x = age, y = .pred, colour = treatment),
    linewidth = 1.3
  ) +
  
  labs(
    title = "Interaction Model: AF vs Age",
    x = "Age",
    y = "Predicted AF",
    colour = "Treatment"
  ) +
  
  theme_minimal()

#check assumptions
# Extract underlying lm object
interact_engine <- fit_interact |> 
  extract_fit_engine()

gglm_inter <- gglm(interact_engine)

```

```{r}
# =========================
# MODEL COMPARISON TABLE
# =========================

# Extract lm objects
lm_simple <- fit_simple |> 
  extract_fit_engine()

lm_interact <- fit_interact |> 
  extract_fit_engine()

# Build comparison table
model_table <- tibble(
  Model = c(
    "Simple Linear Regression",
    "Interaction Linear Regression"
  ),
  
  AIC = c(
    AIC(lm_simple),
    AIC(lm_interact)
  ),
  
  R_squared = c(
    summary(lm_simple)$r.squared,
    summary(lm_interact)$r.squared
  ),
  
  Adjusted_R_squared = c(
    summary(lm_simple)$adj.r.squared,
    summary(lm_interact)$adj.r.squared
  )
)

# Create gt table
model_gt <- model_table |> 
  gt() |> 
  
  fmt_number(
    columns = c(AIC, R_squared, Adjusted_R_squared),
    decimals = 3
  ) |> 
  
  cols_label(
    Model = "Model",
    AIC = "AIC",
    R_squared = "R²",
    Adjusted_R_squared = "Adjusted R²"
  ) |> 
  
  tab_header(
    title = "Linear Model Comparison",
    subtitle = "Simple vs Interaction Models"
  )
```

## Introduction

The dataset consists of 100 patients and includes demographic characteristics such as age, sex, smoking status, and treatment group, as well as 51 genetic single nucleotide polymorphisms (SNPs). There are two outcome variables: Activating Factor (AF), a continuous measure of response, and Positive Response (PR), a binary indicator of treatment effectiveness.

The dataset was obtained from a collaborator and provided via email.

The key variables of interest were:

-   Treatment: a class; A, B or C

-   Age: the age in years of the participant, ranging from 25 to 45

-   SNPs: a set of 51 genetic predictors, recorded as categorical genotype values

The key research question was to assess the effect of treatment group on AF and the probability of a PR, while accounting for demographic characteristics (age, sex, smoking status) and genetic variation (SNPs).

## Methods

```{r}
#| label: fig-simple-diagnostics
#| fig-cap: "Diagnostic plots for the simple linear regression model predicting Activating Factor (AF) from centred age and treatment. These plots are used to assess model assumptions, including residual patterns, normality, and influential observations."

gglm_simp_lm
```

```{r}
#| label: fig-predicted-actual-comparison
#| fig-cap: "Comparison of predicted and actual Activating Factor (AF) values for the simple and interaction linear regression models. Points closer to the dashed 1:1 line indicate better predictive agreement between observed and model-predicted AF."

Plot_pred_vs_actual <- pred_vs_actual_simp_lr / pred_vs_actual_inter
Plot_pred_vs_actual
```

```{r}
#| label: fig-af-age-comparison
#| fig-cap: "Predicted Activating Factor (AF) across age for the simple and interaction linear regression models. The interaction model allows the relationship between age and AF to differ by treatment group."

Plot_AF_vs_age <- AF_vs_age_simp_lm / AF_vs_age_inter
Plot_AF_vs_age
```

```{r}
#| label: fig-interaction-diagnostics
#| fig-cap: "Diagnostic plots for the interaction linear regression model predicting Activating Factor (AF) from centred age, treatment, and age-by-treatment interaction terms. These plots are used to assess whether the fitted model satisfies linear regression assumptions."

gglm_inter
```

## Results

```{r}
#| label: tbl-interaction-coefficients
#| tbl-cap: "Estimated coefficients from a linear interaction model examining the effects of age, centred at its mean, and treatment on Activating Factor (AF). The intercept represents the expected AF for the reference group, Treatment A, at the mean age. Treatments B and C are associated with large increases in AF relative to Treatment A at the mean age. Age has a positive effect on AF, and the significant interaction terms indicate that the effect of age is stronger under Treatments B and C compared to Treatment A. All predictors are highly statistically significant ($p < 0.001$)."

coef_table <- fit_interact |> 
  extract_fit_parsnip() |> 
  tidy() |> 
  
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3)
  ) |> 
  
  gt() |> 
  
  fmt_scientific(
    columns = p.value,
    decimals = 2
  ) |> 
  
  tab_header(
    title = "Linear Interaction Model"
  )

coef_table
```

```{r}
#| label: fig-interaction-model
#| fig-cap: "Interaction model results for AF. Predicted values closely match observed values in the top panel, indicating strong model fit. AF increases with age across all treatments in the bottom panel, with larger increases under Treatments B and C, demonstrating an age-by-treatment interaction."

inter_visual <- pred_vs_actual_inter / AF_vs_age_inter

inter_visual
```

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Comparison of simple and interaction linear regression models predicting Activating Factor (AF) using centred age and treatment. The interaction model, which includes age-by-treatment terms, demonstrates improved model fit, with a lower AIC and higher $R^2$ and adjusted $R^2$ values compared to the simple model. This indicates that allowing the effect of age to vary across treatment groups provides a better explanation of variation in AF."

model_gt
```

## Discussion

## Appendix

```{r, eval = FALSE, include = FALSE}
ggsave(filename = here::here("figs", "gglm_simp_lm.png"),
  plot = gglm_simp_lm,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(filename = here::here("figs", "pred_vs_actual_comparison.png"),
  plot = Plot_pred_vs_actual,
  width = 12,
  height = 10,
  dpi = 300
)

ggsave(filename = here::here("figs", "AF_vs_age_comparison.png"),
  plot = Plot_AF_vs_age,
  width = 12,
  height = 10,
  dpi = 300
)

ggsave(filename = here::here("figs", "gglm_inter.png"),
  plot = gglm_inter,
  width = 8,
  height = 6,
  dpi = 300
)

gtsave(coef_table, here::here("figs", "coef_table.png"))

gtsave(model_gt, here::here("figs", "model_gt.png"))

ggsave(filename = here::here("figs", "inter_visual.png"),
  plot = inter_visual,
  width = 8,
  height = 6,
  dpi = 300
)
```
